\section{Dictionaries}\label{sec:Dictionaries}

\begin{figure}
  \centering
\begin{verbatim}
aAlf  = "Alf"   -- entire alphabet
aObs  = "Obs"   -- all undashed variables
aObs' = "Obs'"  -- all dashed variables
aMdl  = "Mdl"   -- all undashed model variables
aMdl' = "Mdl'"  -- all dashed model variables
aScr  = "Scr"   -- all undashed script variables
aScr' = "Scr'"  -- all dashed script variables
aDyn  = "Dyn"   -- all undashed dynamic observables
aDyn' = "Dyn'"  -- all dashed dynamic observables
aStc  = "Stc"   -- all undashed static parameters
laws  = "laws"  -- all available laws
\end{verbatim}
  \caption{Reserved Dictionary Names}\label{fig:rsvd-names}
\end{figure}

The dictionary datatype (\texttt{Dict s}), 
parameterised with a generic type \texttt{s},
is critical to the functioning of the calculator.
\begin{verbatim}
type Dict s = M.Map String (Entry s)
-- M is the renamed import of Data.Map
\end{verbatim}
It is the basic way in which the user of calculator
describes the alphabet, definitions and laws associated with their theory.

The dictionary uses the Haskell \texttt{String} datatype for keys,
and contains four different kinds of entries: alphabets,
expressions, predicates and laws.
{\small
\begin{verbatim}
data Entry s = AlfEntry .. | ExprEntry .. | PredEntry .. | LawEntry ..
\end{verbatim}
}
For simplicity, there is only one namespace involved,
and some names are reserved for special purposes.
These are listed in Fig. \ref{fig:rsvd-names},
along with Haskell identifiers for them.

There are ten names that describe different
parts of the theory alphabet.
While is is possible to define these individually,
this can be quite error-prone,
so a function is provided to construct all these entries
from three basic pieces of information:
program variable names (`script');
auxiliary variable names (`model',
e.g. variables like $ok$ or $tr$ that don't represent variable values,
but instead some other observable program property of interest);
and static parameter vairiable names. 
\begin{verbatim}
stdAlfDictGen :: [String] -> [String] -> [String] -> Dict s
\end{verbatim}
All lists contain undashed names, with dashes added were required
by the function.

So, the alphabet entries for the UTCP theory are defined as follows:
\begin{code}
dictAlpha = stdAlfDictGen ["s"] ["ls"] ["g","in","out"]
\end{code}
All of these entries will be of kind \texttt{AlfEntry},
i.e, just lists of the relevant variables.
\begin{verbatim}
   AlfEntry { avars   :: [String]}
\end{verbatim}

There are two important utility functions,
one that builds dictionaries from lists of string/entry pairs,
and another that merges two dictionaries, resolving conflicts
by merging entries if possible, other wise favouring the
second dictionary:
\begin{verbatim}
makeDict :: [(String, Entry s)] -> Dict s
dictMrg :: Dict s -> Dict s -> Dict s
\end{verbatim}