\section{Laws}\label{sec:Laws}

In addition to the global simplifier and definition expansion facility,
we have three broad classes of laws that can be invoked from the REPL:
Reduce; Conditional Reduce; and Loop Unroll.

The way the latter three laws are applied is somewhat different
to the behaviour of either the simplifier or definition expansion.
Instead the reserved dictionary key \verb$"laws"$ is used
to lookup a special dictionary entry
\begin{verbatim}
 | LawEntry {  -- about useful laws
     reduce  :: [RWFun s]            -- reduction laws
   , creduce :: [CRWFun s]           -- conditional reductions
   , unroll  :: [String -> RWFun s]  -- loop unrollers }
\end{verbatim}

\subsection{Reduce}

The reduce entry is a list of \texttt{RWFun},
which are defined as follows:
\begin{verbatim}
type RWFun s = Dict s     -- dictionary
            -> Pred s     -- invariant predicate
            -> Pred s     -- target predicate
            -> RWResult s
\end{verbatim}
The first predicate argument is used to supply an invariant
assertion for those reduction rules that require one.
It is a recent new feature of the calculator, 
not required for this UTCP theory,
and its use is beyond the scope of this paper.



When asked to do a reduce,
the calculator then does a top-down, left-to-right search,
where at each point it tries all the laws in its reduce list, in order,
with the current composite being passed in as the second predicate
argument.
It terminates at the point of first success (a non-\texttt{Nothing} outcome).
A reduce law is an equation of the form $P = Q$,
where we search for instances of $P$ and replace them with the
corresponding instance of $Q$.
The idea is that we pattern-match on predicate syntax
with the second predicate argument,
to see if a law is applicable (we have its lefthand-side),
and if so,
we then build an appropriate instance of the righthand-side.
The plan is that we gather all these pattern/outcome pairs
in one function definition,
which will try them in order.
This is in direct correspondence with Haskell pattern-matching.
So for UTCP we have a function called \texttt{reduceUTCP},
structured as follows:
\begin{verbatim}
reduceUTCP d inv (...1st law pattern...) = 1st-outcome
reduceUTCP d inv (...2nd law pattern...) = 2nd-outcome
...
reduceUTCP _ _ _ = Nothing  -- catch-all at end, no change
\end{verbatim}
The last clause matches any predicate.
and simply returns it with a null string,
indicating no change took place.

A simple example of such a pattern is the following encoding
of $\Skip;P = P$ :
\begin{code}
reduceUTCP d inv (Comp "Seq" [(Comp "Skip" []), pr])
   =  Just ( ";-lunit", pr, True )
\end{code}
The pattern
matches a composite called ``Seq'',
with a argument list containing two predicates.
The first predicate pattern
matches a ``Skip'' composite with no further sub-arguments.
The second argument pattern matches an arbitrary predicate
($P$ in the law above).
The righthand-side constructs a \texttt{RWResult} return value,
with the string being a justification note that says a reduction-step
using a law called ``$;$-lunit'' was applied,
and noting that the top-level composite (the ``Seq'') was modified.


\subsection{Conditional Reduce}

A \texttt{CRWResult} is a \texttt{RWResult} that has been adapted,
so that instead of returning one result if successful,
it returns a list of possible results,
each paired with a side-condition predicate.
\begin{verbatim}
type CRWResult s
 = Maybe ( String      -- justification
         , [( Pred s   -- side-condition to be discharged
            , Pred s   -- modified predicate
            , Bool)])  -- True if top-level modified
type CRWFun s = Dict s -> Pred s -> CRWResult s
\end{verbatim}
A conditional reduce law is an equation as per reduce,
but with conditional outcomes, e.g. $P = Q_1 \cond{C} Q_2$.
Matching an instance of $P$ will return a list of two pairs,
the first being $(C,Q_1)$, the second $(\lnot C,Q_2)$.
No attempt is made to evaluate $C$, but instead the REPL
asks the user to choose.
This is a key design decision for the calculator.
A general purpose predicate evaluator requires implementing
lots of theories about numbers, sets, lists,
and whatever else might be present.
Given the scope and purpose of this calculator is is
much more effective to let the user choose.

For an example,
here is one pattern of the conditional reduce
function for UTCP.
Given $\vec x$ a list of unique variables,
and $\vec e$ a list of the same length of expressions,
with $\vec x \subseteq \setof{s,ls}$
we have:
\RLEQNS{
   c[\vec e/\vec x]
   &\implies&
   (c * P)[\vec e/\vec x] = P[\vec e/\vec x] ; c * P
\\ \lnot c[\vec e/\vec x]
   &\implies&
   (c * P)[\vec e/\vec x] = \Skip[\vec e/\vec x]
}
\begin{code}
creduceUTCP d (PSub w@(Comp "Iter" [c,p]) sub)
 | isCondition c && beforeSub d sub
 = Just( "loop-substn", [ctrue,cfalse] )
 where
   csub = PSub c sub
   ctrue  = (       csub, mkSeq (PSub p sub) w, diff )
   cfalse = ( mkNot csub, PSub mkSkip sub, diff )
\end{code}
Here \texttt{mkSeq}, \texttt{mkNot} and \texttt{mkSkip} build sequential composition,
negations and standard UTP skip ($\Skip$) respectively.



\subsection{Loop Unroll}


Iteration is typically defined in UTP as the least fixed point
w.r.t to the refinement ordering
that also involves sequential composition,
which itself is defined using existential quantification,
and $\Skip$.
\RLEQNS{
   c * P &\defs& \mu L @ (P ; L) \cond c \Skip
\\ P ; Q
   &\defs&
   \exists s_m,ls_m @
     P[s_m,ls_m/s',ls']
     \land
     Q[s_m,ls_m/s,ls]
\\ \Skip &\defs& s'=s \land ls'=ls
}
We do not want to explicitly handle quantifiers,
or fixed-points.
Instead we prefer to use the loop unrolling law,
as this is much more useful for the kinds of calculations we encounter.
\RLEQNS{
   c * P &=& (P \seq c * P ) \cond c \Skip
}
Even more useful are ones that split the conditional
and unroll a number of times ($\seq$ binds tighter than $\lor$
but looser than $\land$):
\RLEQNS{
   c * P &=& \lnot c \land \Skip \lor c \land P \seq c * P
\\ &=& \lnot c \land \Skip
       \lor c \land P \seq c \land \Skip
       \lor c \land P \seq c * P
\\ &=& \lnot c \land \Skip
       \lor c \land P \seq c \land P \seq c \land \Skip
       \lor c \land P \seq c \land P \seq c * P
\\ &=& \ldots
}
The loop unroll functions are like those for reduce
but have an extra string argument.
\begin{verbatim}
    unroll  :: [String -> RWFun s]
\end{verbatim}
When the user enters a command of the form \verb@"lsss"@,
the loop unroll facility is activated,
and the string \verb@"sss"@ is passed as the first argument to the functions
above. It is up to the user to decide how to interpret
these strings---but the most useful is to treat them as specifying
the number of unrollings to do.
We won't give an example here of the use of unrolling.
%\begin{code}
%wUnroll :: Ord s => String -> DictRWFun s
%wUnroll ns d mw@(_,Comp nm [mpr])
% | nm == nW = ( "W-unroll" ++ ntag ns, wunroll n )
% where
%
%   ntag "" = ""
%   ntag ns = '.':ns
%
%   n | null ns = 0
%     | isDigit $ head ns = digitToInt $ head ns
%     | otherwise = 0
%
%   wunroll 0  =  bCond (bSeq mpr mw) (bNot lsout) bSkip
%   wunroll 1  =  bOr [ loopdone
%                     , bSeq (loopstep 1) mw]
%   wunroll 2  =  bOr [ loopdone
%                     , bSeq (loopstep 1) loopdone
%                     , bSeq (loopstep 2) mw]
%   wunroll 3  =  bOr [ loopdone
%                     , bSeq (loopstep 1) loopdone
%                     , bSeq (loopstep 2) loopdone
%                     , bSeq (loopstep 3) mw]
%   wunroll _  =  bCond (bSeq mpr mw) (bNot lsout) bSkip
%
%   loopdone = bD out
%   loopstep 1 = bAnd [bM out, mpr]
%   loopstep n = bSeq (loopstep (n-1)) (loopstep 1)
%
%wUnroll _ _ mpr = ( "", mpr )
%\end{code}
%It does up to 3 unrollings---we won't explain all the various bits
%but just point out that \texttt{loopdone} is $\lnot c \land \Skip$
%(under the hood) and \texttt{loopstep} n is $(c \land P)^n$.


\subsection{Bringing it all together}

We make these two reduction functions ``known'' to the calculator
by adding them into a dictionary.
\begin{code}
lawsUTCPDict
 = makeDict [("laws", LawEntry [reduceUTCP] [creduceUTCP] [])]
\end{code}
We then can take a number of partial dictionaries and use various
dictionary functions,
defined in \texttt{CalcPredicates}, to merge them together.
\begin{code}
dictUTCP = foldl1 dictMrg [ alfUTCPDict, ..., lawsUTCPDict]
\end{code}
The main method of working with dictionaries 
is to construct small ones
focussed on some specific area of interest.
These can then be combined in different ways to provide a number of complete
dictionaries that can vary in the order in which things are tried.









%\subsection{UTCP Recognisers}
%
%Some laws require matching that is a bit more sophisticated.
%For example,
%consider a useful reduction for tidying up at the end,
%assuming that $ls' \notin A$ and $ls \notin B$, and both $k$ and $h$
%are ground:
%\[
%   (A \land ls'=k) ; (B \land ls'= h)
%   \quad\equiv\quad
%   (A;B) \land ls'=h
%   \qquad \elabel{$ls'$-cleanup}
%\]
%However, we want this law to work when both $A$
%and $B$ are themselves conjunctions, with the $ls'=\dots$
%as part of the same conjunction, located at some arbitrary position.
%We can break the problem into two parts.
%First we do a top-level pattern match
%to see that we have a top-level sequential composition
%of two conjunctions,
%then we use a function that will check both conjunction predicate-lists
%for the existence of a $ls'=\dots$ component,
%and that everything else also satisfies the requirements regarding
%the occurrence, or otherwise of $ls$ or $ls'$:
%\begin{code}
%reduceUTCP d pr@(_,Comp "Seq" [ (_,Comp "And" pAs)
%                              , (_,Comp "And" pBs)])
% = case isSafeLSDash d ls' ls' pAs of -- no ls' in rest of pAs
%    Nothing -> lred "" pr
%    Just (_,restA) ->
%     case isSafeLSDash d ls' ls pBs of -- no ls in rest of pBs
%      Nothing -> lred "" pr
%      Just (eqB,restB)
%       -> lred "ls'-cleanup" $   -- build RHS
%             comp "And" [ comp "Seq" [ bAnd restA
%                                     , bAnd restB ]
%                        , eqB ]
%\end{code}
%The function \texttt{isSafeLSDash}
%is designed to
%(i) locate the $ls'=e$ conjunct and check that its rhs is a ground expression;
%(ii) check that none of the remaining conjuncts make use of the
%`unwanted' version of the label-set variable ($ls$ or $ls'$);
%and (iii), if all ok, return a pair
%whose first component is the ($ls'=\dots$) equality,
%and whose second is the list of other conjuncts.
%To achieve (i) above,
%we make use of two functions provided by the \texttt{CalcRecogniser} module:
%\begin{verbatim}
%mtchNmdObsEqToConst :: Ord s => String -> Dict s -> Recogniser s
%matchRecog :: (Ord s, Show s)
%           => Recogniser s -> [MPred s]
%           -> Maybe ([MPred s],(MPred s,[MPred s]),[MPred s])
%\end{verbatim}
%where
%\begin{verbatim}
%type Recogniser s = MPred s -> (Bool, [MPred s])
%\end{verbatim}
%A recogniser is a function that takes a predicate
%and if it ``recognises'' it, returns \texttt{(True, parts)},
%where parts are the subcomponents of the predicate in some order.
%The recogniser \texttt{mtchNmdObsEqtoConst v d} matches a predicate of the form
%\texttt{Equal (Var v) k}, returning a list with both parts.
%The function \texttt{matchRecog} takes a recogniser and list of predicates
%and looks in the list for the first predicate to satisfy
%the recogniser, returning a triple of the form
%(before,satisyingPred,after).
%If the recogniser succeeds,
%we then check the validity of the expression,
%and the absence of the unwanted variable from the
%rest of the conjuncts --- using boolean function
% \texttt{dftlyNotInP} (``definitely not in $P$''),
% so handling task (ii) above.
%\begin{code}
%isSafeLSDash d theLS unwanted prs
% = case matchRecog (mtchNmdObsEqToConst theLS d) prs of
%    Nothing -> Nothing
%    Just (pre,(eq@(_,Equal _ k),_),post) ->
%     if notGround d k
%      then Nothing
%      else if all (dftlyNotInP d unwanted) rest
%       then Just (eq,rest)
%       else Nothing
%     where rest = pre++post
%\end{code}
